\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
got it down to eight pages using negative vertical space
made text chars more consistent
Josh conclusion

Viewed as a probabilistic inference problem,
our goal is to infer the DSL (written $\mathcal{D}$)
along with a program (written $p$) for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.

@@

\subsection{Symbolic Regression}\label{regressionSection}
We show how to use \system to infer programs containing both discrete
structure and continuous parameters. The high-level idea is to write programs with unspecified real-valued parameters, and to fit those parameters by differentiating through the program
and optimizing using gradient descent.
We have the algorithm 
solve a set of 200 symbolic regression problems, each a polynomial of
degree 1 to 4, where our observations $x$ are 10
input/output examples.%, which we write as $x = \left\{(i_n,o_n)
%\right\}_{n\leq N}$. %% For example, one task is to infer a program
%% calculating $3x + 2$, and the observations are the input-output
%% examples $\left\{(-1,-1),(0,2),(1,5) \right\}$.

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom.
For example, it represents the linear function $-x+2$ with the program
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} which has two continuous degrees of freedom, and represents quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}
which has five continuous degrees of freedom.
This arises from our Bayesian framing -- both the bias towards shorter programs and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some learned DSL primitives for symbolic regression. The system starts with addition, multiplication, and real numbers, and learns to build polynomials up to 4th order.}\label{regressionDSL}
\end{table}

@@

  \\\midrule\multicolumn{6}{c}{\emph{List functions (rich DSL)}}\\\midrule
  \% solved&\textbf{86\%} &84\% &60\%&74\%&70\%\\
  Solve time&  0.8s&0.7s&1.0s&1.0s&1.1s
  \\\bottomrule
  \end{tabular}
\caption{\% solved w/ 5 sec timeout. Solve time: averaged over solved
  tasks.  RF/DC: trained like RobustFill/DeepCoder.  PCFG: model w/o
  structure learning.  Enum: model w/o any learning.  MDL:
  $-\expect\left[\probability[x|p] \right]$. For domains other than
  symbolic regression MDL is 0 nats. For symbolic regression MDL is a
  proxy for \# of continuous parameters.}\label{baselineComparisons} \end{table}

@@

 \noindent \emph{Inventing new subroutines for program induction:}
 Several program induction algorithms, most prominently the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}, take as their goal to learn new, reusable subroutines that are shared in a multitask setting. We find this work inspiring and motivating,
 and extend it along two dimensions: (1) we propose a new algorithm for
 inducing reusable subroutines, based on Fragment Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing};
 and (2) we show how to combine these techniques with bottom-up neural recognition models.
 Other instances of this related idea are~\cite{DBLP:conf/icml/LiangJK10}, Schmidhuber's OOPS model~\cite{schmidhuber2004optimal}, and predicate invention in ILP~\cite{DBLP:conf/ecai/LinDETM14}.

@@

 \section{Contribution and Outlook}
 We contributed a model, \systemEnding, that learns to program from examples by
 bootstrapping a DSL with new domain-specific primitives that the algorithm itself discovers.
  Alongside this DSL, we train a bottom-up
 neural recognition model that learns how to efficiently deploy the
 DSL and hone in on the right programs.  We believe that this
 integration of top-down symbolic representation and bottom-up neural
 networks -- both of them learned -- is a useful general approach for
 getting program induction off the ground.

>>>>

Viewed as a probabilistic inference problem,
our goal is to jointly infer a single DSL (written $\mathcal{D}$)
and many programs (written $p$), one for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.

@@

\subsection{Symbolic Regression}\label{regressionSection}
We show how to use \system to infer programs containing both discrete
structure and continuous parameters. The high-level idea is to write programs with unspecified real-valued parameters, and to fit those parameters by differentiating through the program
and optimizing using gradient descent.
We task our model
with solving 200 symbolic regression problems, each a polynomial of
degree 1 to 4, where our observations $x$ are 10
input/output examples.%, which we write as $x = \left\{(i_n,o_n)
%\right\}_{n\leq N}$. %% For example, one task is to infer a program
%% calculating $3x + 2$, and the observations are the input-output
%% examples $\left\{(-1,-1),(0,2),(1,5) \right\}$.

@@

\system learns a DSL containing templates for polynomials of different orders,
which lets the algorithm quickly find the functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
We also found that the algorithm discovers programs that minimize the number of continuous degrees of freedom.
For example, it represents the linear function $-x+2$ with the program
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} which has two continuous degrees of freedom, and represents quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing -- both the bias towards shorter programs and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some learned DSL primitives for symbolic regression. The system starts with addition, multiplication, and real numbers, and learns to build polynomials up to 4th order.}\label{regressionDSL}
\end{table}

@@

  \\\midrule\multicolumn{6}{c}{\emph{List functions (rich DSL)}}\\\midrule
  \% solved&\textbf{86\%} &84\% &60\%&74\%&70\%\\
  Solve time&  0.8s&0.7s&1.0s&1.0s&1.1s
  \\\bottomrule
  \end{tabular}
\caption{\% solved w/ 5 sec timeout. Solve time: averaged over solved
  tasks.  RF/DC: trained like RobustFill/DeepCoder.  PCFG: model w/o
  structure learning.  Enum: model w/o any learning.  MDL:
  $-\expect\left[\probability[x|p] \right]$. For domains other than
  symbolic regression MDL is 0 nats. For symbolic regression MDL is a
  proxy for \# of continuous parameters.}\vspace{-0.5cm}\label{baselineComparisons} \end{table}

@@

 \noindent \emph{Inventing new subroutines for program induction:}
 Several program induction algorithms, most prominently the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}, take as their goal to learn new, reusable subroutines that are shared in a multitask setting. We find this work inspiring and motivating,
 and extend it along two dimensions: (1) we propose a new algorithm for
 inducing reusable subroutines, based on Fragment Grammars~\cite{tim};
 and (2) we show how to combine these techniques with bottom-up neural recognition models.
 Other instances of this related idea are~\cite{DBLP:conf/icml/LiangJK10}, Schmidhuber's OOPS model~\cite{schmidhuber2004optimal}, and predicate invention in ILP~\cite{DBLP:conf/ecai/LinDETM14}.

@@

 \section{Contribution and Outlook}
 We contributed a model, \systemEnding, that learns to program by
 bootstrapping a DSL with new domain-specific primitives that the algorithm itself discovers.
  Alongside this DSL, we train a bottom-up
 neural recognition model that learns how to efficiently deploy the
 DSL and hone in on the right programs.
 %% We believe that this
 %% integration of top-down symbolic representation and bottom-up neural
 %% networks -- both of them learned -- is a useful general approach for
 %% getting program induction off the ground. 
Many directions remain open for future work. Two immediate goals are to
integrate more sophisticated neural recognition models~\cite{devlin2017robustfill} and
program synthesizers~\cite{solar2008program}, which may improve performance in some
domains over the simple, generic methods used here, and to explore DSL
meta-learning: Can we find a \emph{single} universal primitive set that could
effectively bootstrap DSLs for new domains, including the four domains
considered but many others?
